---
title: "Backend provenance and CUDA diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Backend provenance and CUDA diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`cudalearnr` uses one API on CPU and CUDA, but it does not treat a CUDA-aware
function name as proof that every stage ran on a GPU. Version 0.2.0 records the
device request and actual execution of every stage.

This tutorial first runs entirely on the CPU. The CUDA section is optional and
executes only when diagnostics prove that a usable device exists.

## A reproducible CPU workflow

Use `device = "cpu"` when you need a portable, guaranteed CPU run. This path
does not inspect or initialize the optional CUDA runtime.

```{r cpu-data}
library(cudalearnr)

x <- scale(as.matrix(iris[, 1:4]))
rownames(x) <- paste0("flower_", seq_len(nrow(x)))

decomposition <- cuda_svd(
  x,
  nu = 2,
  nv = 2,
  device = "cpu"
)
pca <- cuda_pca(
  x,
  n_components = 2,
  device = "cpu"
)
distance <- cuda_distance(
  pca$x,
  device = "cpu"
)
neighbors <- cuda_knn(
  pca$x,
  k = 5,
  batch_size = 32,
  device = "cpu"
)
clusters <- cuda_kmeans(
  pca$x,
  centers = 3,
  seed = 1,
  device = "cpu"
)
```

The numerical results remain ordinary R objects. Provenance is separate,
stable metadata:

```{r cpu-provenance}
cuda_provenance(decomposition)
cuda_provenance(pca)
cuda_provenance(distance)
cuda_provenance(neighbors)
cuda_provenance(clusters)
```

`cuda_provenance()` returns one row per stage. Its columns have deliberately
different meanings:

| Column | Meaning |
|---|---|
| `requested_device` | What selected the stage: `"cpu"`, `"cuda"`, `"auto"`, a fixed CPU algorithm, or an inherited choice. |
| `device` | Where that stage actually computed: `"cpu"` or `"cuda"`. |
| `backend` | The concrete implementation, such as `base`, `stats`, or `torch`. |
| `selection_reason` | Why the device was selected, including explicit requests, fixed CPU work, input transfers, or CUDA unavailability. |
| `fallback` | `TRUE` only when an `"auto"` request selected CPU because CUDA was unavailable. |
| `output_device` | Where the stage result was materialized. Returned R matrices and lists normally live on CPU even after CUDA computation. |

The table has a `compute_device` attribute. It is `"cpu"` when all stages are
CPU, `"cuda"` when all compute stages are CUDA, and `"hybrid"` when both occur:

```{r aggregate-device}
pca_provenance <- cuda_provenance(pca)
attr(pca_provenance, "compute_device")
```

An output on CPU is not evidence of a fallback. CUDA kernels commonly return
an R matrix on CPU. Conversely, `fallback = TRUE` records a device-selection
decision; it is not used to hide a CUDA execution error.

## Stage-by-stage backend semantics

The following table describes successful CPU and explicit CUDA requests.

| Function | Stage | CPU backend | CUDA backend | Output |
|---|---|---|---|---|
| `cuda_svd()` | `decomposition` | `base` | `torch` | CPU R list |
| `cuda_pca()` | `preprocessing` | `stats` | `torch` | selected device |
| `cuda_pca()` | `decomposition` | `stats` | `torch` | CPU R list |
| `cuda_distance()` | `distance` | `base` | `torch` | CPU dense matrix |
| `cuda_knn()` | `distance` | `base` | `torch` | CPU distance block |
| `cuda_knn()` | `neighbor_selection` | `base` | `base` | CPU R list |
| `cuda_kmeans()` | `initialization` | `base` | `base` | CPU |
| `cuda_kmeans()` | `distance` | `base` | `torch` | CPU distance matrix |
| `cuda_kmeans()` | `assignment` | `base` | `base` | CPU |
| `cuda_kmeans()` | `center_update` | `base` | `base` | CPU R list |

Thus SVD, PCA, and distance have a CUDA aggregate when explicitly run on CUDA.
kNN and k-means are intentionally hybrid: CUDA accelerates distance
calculation, while deterministic selection, assignments, and centre updates
remain in R on the CPU.

If a CUDA `cudatensor` must first be materialized as an R matrix, provenance
also includes an `input_materialization` stage. This makes a device transfer
visible instead of presenting the subsequent kernel as the whole workflow.

## Runtime diagnostics and device requests

Diagnostics are non-destructive: they inspect the optional torch runtime but
never install or download it.

```{r diagnostics}
diagnostics <- cudatensr::cuda_diagnostics()
diagnostics
```

Device requests follow these rules:

- `"cpu"` is explicit and always uses the portable backend.
- `"cuda"` is strict. If CUDA is unavailable, the call raises a
  `cudaverse_cuda_unavailable` error; it does not fall back.
- `"auto"` uses CUDA when diagnostics report a usable device. Otherwise it
  selects CPU and records `fallback = TRUE` plus a stable reason such as
  `torch_not_installed` or `cuda_unavailable`.
- Once CUDA has been selected, an execution error is reported as an error. It
  is not retried silently on CPU.

## Optional CUDA run

The same data can be used for a small CUDA smoke test. The conditional keeps
the vignette runnable on machines without a GPU.

```{r optional-cuda}
if (isTRUE(diagnostics$cuda_available)) {
  cuda_pca_fit <- cuda_pca(
    x,
    n_components = 2,
    device = "cuda"
  )
  cuda_neighbors <- cuda_knn(
    cuda_pca_fit$x,
    k = 5,
    batch_size = 32,
    device = "cuda"
  )

  cuda_provenance(cuda_pca_fit)
  cuda_provenance(cuda_neighbors)
} else {
  message("CUDA example skipped: ", diagnostics$reason)
}
```

For kNN, a CUDA distance stage followed by CPU neighbour selection produces
`compute_device = "hybrid"`. This is expected and is stronger evidence than
checking only the legacy `$device` field.

## Dense-memory boundaries

`cuda_pca()` accepts a dense observation-by-feature matrix. A double matrix
alone uses roughly `8 * nrow(x) * ncol(x)` bytes, and centring, scaling, SVD,
and CPU/GPU copies require additional working memory. Reduce the number of
observations or features before PCA when the dense representation is too
large.

`cuda_distance()` intentionally returns the complete dense pairwise matrix,
requiring about `8 * nrow(x) * nrow(y)` bytes for doubles.

`cuda_knn()` is exact and quadratic in time, but `batch_size` bounds its
resident distance block to approximately
`8 * min(batch_size, nrow(x)) * nrow(x)` bytes. Its returned distance matrix
uses about `8 * nrow(x) * k` bytes, in addition to the integer index matrix.
Lower `batch_size` to reduce peak memory; it does not change selected
neighbours.

## Hardware-enforced parity gate

Ordinary package checks use portable CPU paths and may conditionally skip GPU
examples. That is not accepted as CUDA coverage. The cudaverse hardware
workflow runs on a self-hosted runner labelled `cuda`, sets
`CUDAVERSE_REQUIRE_CUDA=true`, proves NVIDIA visibility with `nvidia-smi`, and
requires both `torch::cuda_is_available()` and a positive CUDA device count.
It then compares the public CPU and CUDA paths and verifies their provenance.

Projects reusing the same convention can turn the environment marker into a
hard local gate:

```{r hardware-gate}
require_cuda <- identical(
  tolower(Sys.getenv("CUDAVERSE_REQUIRE_CUDA", unset = "false")),
  "true"
)
if (require_cuda && !isTRUE(diagnostics$cuda_available)) {
  stop("CUDA hardware is required, but no usable device was detected.")
}
```

The package-level CUDA workflow runs manually, or automatically when the
repository variable `CUDAVERSE_NVIDIA_CI` is set to `enabled`. A missing or
broken GPU on that required job is a failure, never a successful skip.
